---
title: "Dimensionality reduction"
---
  
```{r load_modules, echo=FALSE, include=FALSE}
library(Seurat)
# library(RColorBrewer)
# library(umap)
```

# Define settings

Define I/O
```{r define_io, echo=FALSE}
# Load default settings
source("/Users/ricard/10x_gastrulation_TetChimera/settings.R")

# io$outdir <- paste0(io$basedir,"/results/dimensionality_reduction")
```

Define options
```{r}
opts$batches <- c(
	"E75_TET_TKO_L002",
	"E75_WT_Host_L001"
	# "E85_Rep1_TET_TKO_L004",
	# "E85_Rep2_TET_TKO_L006",
	# "E85_Rep1_WT_Host_L003",
	# "E85_Rep2_WT_Host_L005"
	# "E125_DNMT3A_HET_A_L001",
	# "E125_DNMT3A_HET_A_L003",
	# "E125_DNMT3A_KO_B_L002",
	# "E125_DNMT3A_KO_E_L004"
)

# opts$target <- c("TKO","WT")
```

Update sample metadata
```{r load_metadata, echo=FALSE}
sample_metadata <- sample_metadata %>%
  .[batch%in%opts$batches & pass_QC==T]
table(sample_metadata$batch)
```

Load cell type annotations
```{r}
foo <- fread("/Users/ricard/data/10x_gastrulation_TetChimera/processed/second_batch/mapping/mnn/mapping_mnn_WT.txt.gz") %>%
  .[class=="WT"]
bar <- fread("/Users/ricard/data/10x_gastrulation_TetChimera/processed/second_batch/mapping/mnn/mapping_mnn_TET_TKO.txt.gz") %>%
  .[class=="TET_TKO"]

celltypes <- rbind(foo,bar) %>%
  .[cell%in%sample_metadata$cell] %>%
  .[,c("cell","celltype.mapped","stage.mapped","celltype.score","cellstage.score")]

sample_metadata <- sample_metadata %>% merge(celltypes)
```

# Load data

Load RNA expression data as Seurat object
```{r load_data, echo=FALSE}
seurat <- readRDS(io$seurat)#[,sample_metadata$cell]
dim(seurat)
```

```{r}
foo <- sample_metadata %>% as.data.frame %>% tibble::column_to_rownames("cell") %>%
  .[colnames(seurat),]

stopifnot(colnames(seurat) == rownames(foo))
# seurat <- AddMetaData(seurat, foo)  # not working????
seurat@meta.data <- foo
```

Remove genes that are not expressed
```{r}
foo <- Matrix::rowMeans(seurat@assays$RNA@counts)
seurat <- subset(seurat, features = names(which(foo>1e-6)))
```

# Parse data 

Normalise
```{r}
# SCTransform
# seurat <- SCTransform(seurat)

# LogNormalize
seurat <- NormalizeData(seurat, normalization.method = "LogNormalize")

# Scale
seurat <- ScaleData(seurat)
```

Change gene names
```{r}
# rownames(sce) <- rowData(sce)$symbol
```

# Select HVG

Option 1, using FindVariableFeatures
```{r}
seurat <- FindVariableFeatures(seurat, selection.method = 'vst', nfeatures = 1000)
# length(seurat@assays$RNA@var.features)
```

Option 2, using precomputed marker genes
```{r}
marker_genes.dt <- fread(io$atlas.marker_genes)
hvg <- unique(marker_genes.dt$ens_id)
hvg <- hvg[hvg%in%rownames(seurat)]

seurat@assays$RNA@var.features <- hvg
```

# Dimensionality reduction

PCA 
```{r message=FALSE}
seurat <- RunPCA(seurat, 
  features = VariableFeatures(seurat),
  npcs = 50,
  verbose = FALSE
)
```

Correlation between PCs and batch
```{r}
r <- cor(
  seurat@reductions$pca@cell.embeddings,
  as.numeric(as.factor(seurat@meta.data$batch))
)[,1]
r
```

UMAP
```{r}
seurat <- RunUMAP(seurat, 
  # dims = 1:50, 
  dims = which(r<=0.5), 
  reduction = "pca"
)
```

# Plot 

```{r}
Idents(seurat) <- "batch"
# Idents(seurat) <- "celltype.mapped"
# levels(seurat)
```


```{r}
DimPlot(seurat, label = FALSE, reduction = 'pca', dims=c(10,11))
```

```{r}
DimPlot(seurat, label = FALSE, reduction = 'umap')
DimPlot(seurat, label = FALSE, reduction = 'umap', cols = opts$colors1) + NoLegend()
```

Explore marker genes
```{r}
genes.to.plot <- c("ENSMUSG00000026566")

FeaturePlot(seurat, features = genes.to.plot, reduction='umap')
```

