---
title: "Velocyto analysis"
---

```{r}
library(velocyto.R)
```

## Define settings

Load default settings
```{r}
source("/Users/ricard/10x_gastrulation_TetChimera/settings.R")
```

Define I/O
```{r}
io$loom.dir <- paste0(io$basedir,"/processed/velocyto")
io$outdir <- paste0(io$basedir,"/results/velocyto")
```

Define options
```{r}
# opts$batch
```

## Data loading

Update sample metadata
```{r load_metadata, echo=FALSE}
sample_metadata <- sample_metadata %>%
  .[batch%in%c("E75_WT_Host_L001")]
table(sample_metadata$batch)
```

Load SingleCellExperiment
```{r}
sce <- readRDS(io$sce)[,sample_metadata$cell]
colData(sce) <- sample_metadata %>% as.data.frame %>% tibble::column_to_rownames("cell") %>%
  .[colnames(sce),] %>% DataFrame()
dim(sce)
```


Load loom data
```{r}
ldat <- read.loom.matrices(paste0(io$loom.dir,"/E75_WT_Host.loom"))
# ldat <- read.loom.matrices(sprintf("%s/%s.loom",io$loom.dir,opts$batch))
lapply(ldat,dim)
```

Load gene metadata
```{r}
gene_metadata <- fread(io$gene_metadata) %>% .[,c("ens_id","symbol")] %>%
  .[symbol!="" & ens_id%in%rownames(sce)]
```

# Parse data 

Update cell names to match HDF5 idiosyncrasies
```{r}
colnames(sce) <- colnames(sce) %>% 
  map_chr(~ sub("E75_WT_Host_L001_","E75_WT_Host:",.) %>% substr(1,28) %>% paste0("x")
)

sample_metadata[,cell := sub("E75_WT_Host_L001_","E75_WT_Host:",cell) %>% substr(1,28) %>% paste0("x")]
```

Sanity checks
```{r}
mean(colnames(sce) %in% colnames(ldat$spliced))
mean(colnames(ldat$spliced) %in% colnames(sce))
```

Subset cells from the velocyto matrices
```{r}
ldat$spliced <- ldat$spliced[,colnames(sce)]
ldat$unspliced <- ldat$unspliced[,colnames(sce)]
ldat$ambiguous <- ldat$ambiguous[,colnames(sce)]
lapply(ldat,dim)
```

Subset genes from the velocyto matrices
```{r}
# genes <- rownames(sce)
genes.symbol <- intersect(rownames(ldat$spliced),gene_metadata$symbol)
genes.ensid <- gene_metadata[symbol%in%genes.symbol & ens_id%in%rownames(sce),ens_id]
ldat$spliced <- ldat$spliced[genes.symbol,]
```

Subset genes from the SCE
```{r}
sce <- sce[genes.ensid,]
```

# Dimensionality reduction

Select HVG
```{r}
decomp <- modelGeneVar(sce)
# decomp <- decomp[decomp$mean > 0.01,]
decomp <- decomp[decomp$mean > 0.001,]
hvgs <- rownames(decomp)[decomp$p.value <= 0.01]

# Subset SingleCellExperiment
sce_filt <- sce[hvgs,]
dim(sce_filt)
```


# PCA

Run PCA
```{r}
sce_filt <- runPCA(sce_filt, ntop = 9999, ncomponents = 25)
reducedDim(sce, "PCA") <- reducedDim(sce_filt, "PCA") 
```


Run UMAP
```{r}
set.seed(42)
# sce_filt <- runUMAP(sce_filt, dimred="PCA")
sce_filt <- runUMAP(sce_filt, dimred="PCA", n_neighbors = 35, min_dist = 0.45)
reducedDim(sce, "UMAP") <- reducedDim(sce_filt, "UMAP") 
```

Plot UMAP

```{r}
to.plot <- reducedDim(sce_filt,"UMAP") %>% as.data.table %>% 
  .[,cell:=colnames(sce_filt)] %>%
    merge(sample_metadata, by="cell")

p <- ggplot(to.plot, aes(x=V1, y=V2, fill=celltype.mapped)) +
  geom_point(size=1.5, shape=21, stroke=0.2) +
  # scale_color_manual(values=opts$stage.colors) +
  scale_fill_manual(values=opts$celltype.colors) +
  theme_classic() +
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position="none",
    legend.title=element_blank()
  )

# pdf(paste0(io$outdir,"/rna_umap.pdf"), width=5, height=3, useDingbats = F)
print(p)
# dev.off()
```

## Velocity estimation

Prepare matrices and clustering data:
```{r}
emat <- ldat$spliced
nmat <- ldat$unspliced

# take embedding
emb <- reducedDim(sce_filt,"UMAP")
```

Calculate cell-cell distance matrix using PCA space
```{r}
cell.dist <- dist(reducedDim(sce_filt,"PCA"))
class(cell.dist)
```

```{r}
all(colnames(emat) == colnames(sce))
all(colnames(nmat) == colnames(sce))
```

Filter genes based on the minimum average expresion magnitude (in at least one of the clusters), output total number of resulting valid genes:
```{r}
clusters <- sce$celltype.mapped; names(clusters) <- colnames(sce)

emat <- filter.genes.by.cluster.expression(emat,
  clusters = clusters,
  min.max.cluster.average = 0.5
)

nmat <- filter.genes.by.cluster.expression(nmat,
  clusters = clusters,
  min.max.cluster.average = 0.05
)

length(intersect(rownames(emat),rownames(emat)))
```

Estimate RNA velocity (using gene-relative model with k=20 cell kNN pooling and using top/bottom 2% quantiles for gamma fit):
```{r}
fit.quantile <- 0.02

rvel.cd <- gene.relative.velocity.estimates(
  emat = emat,   # exonic count matric
  nmat = nmat,   # intronic count matric
  deltaT = 1,    # amount of time to project the cell forward
  kCells = 20,   # 	number of k nearest neighbors (NN) to use in slope calculation smoothing
  cell.dist = cell.dist,  # cell distance to use in cell kNN pooling calculations
  fit.quantile = fit.quantile # perform gamma fit on a top/bottom quantiles of expression magnitudes
)
```


Visualize velocity on the embedding, using velocity vector fields:
```{r fig.height=6,fig.width=6}
show.velocity.on.embedding.cor(emb, 
  rvel.cd, 
  n = 300, 
  scale = 'sqrt',
  # cell.colors = ac(cell.colors,alpha = 0.5),
  cex = 0.8,
  arrow.scale = 5,
  show.grid.flow = TRUE,
  min.grid.cell.mass = 0.5,
  grid.n = 40,
  arrow.lwd = 1,
  do.par = F,
  cell.border.alpha = 0.1
)
```


Visualize a fit for a particular gene (we reuse rvel.cd to save on calcualtions here):
```{r fig.height=2.3,fig.width=8}
gene <- "Nfib"
gene.relative.velocity.estimates(emat,nmat,deltaT=1,kCells = 20,kGenes=1,fit.quantile=fit.quantile,cell.emb=emb,cell.colors=cell.colors,cell.dist=cell.dist,show.gene=gene,old.fit=rvel.cd,do.par=T)
```
