---
title: "Lineage validation using marker genes"
---
  
```{r load_modules, echo=FALSE, include=FALSE}
library(Seurat)
library(SingleCellExperiment)
library(scater)
library(scran)
library(ggpubr)
# library(RColorBrewer)
# library(umap)
```

# Define settings

Define I/O
```{r define_io, echo=FALSE}
# Load default settings
source("/Users/ricard/10x_gastrulation_TetChimera/settings.R")

io$outdir <- paste0(io$basedir,"/results/second_batch/lineage_validation")
```

Define options
```{r}
opts$batches <- c(
	# "E75_TET_TKO_L002",
	# "E75_WT_Host_L001"
	"E85_Rep1_TET_TKO_L004",
	"E85_Rep2_TET_TKO_L006",
	"E85_Rep1_WT_Host_L003",
	"E85_Rep2_WT_Host_L005"
	# "E125_DNMT3A_HET_A_L001",
	# "E125_DNMT3A_HET_A_L003",
	# "E125_DNMT3A_KO_B_L002",
	# "E125_DNMT3A_KO_E_L004"
)


opts$stages <- c(
  # "E75"
  "E85"
)

# opts$target <- c("TKO","WT")
```

```{r}
# table(sample_metadata$batch)
# table(sample_metadata$stage)
```

Update sample metadata
```{r load_metadata, echo=FALSE}
sample_metadata <- sample_metadata %>%
  .[batch%in%opts$batches & stage%in%opts$stages & pass_QC==T]

sample_metadata[,celltype.mapped:=stringr::str_replace_all(celltype.mapped,c(" "="_","/"="_"))]
```

```{r}
table(sample_metadata$batch)
table(sample_metadata$stage)
```

```{r}
sample_metadata[,.N,by=c("celltype.mapped","class")] %>%
  dcast(celltype.mapped~class)
```


Load cell type annotations
```{r}
foo <- fread("/Users/ricard/data/10x_gastrulation_TetChimera/results/second_batch/mapping/mnn/mapping_mnn_WT.txt.gz") %>%
  .[class=="WT"]
bar <- fread("/Users/ricard/data/10x_gastrulation_TetChimera/results/second_batch/mapping/mnn/mapping_mnn_TET_TKO.txt.gz") %>%
  .[class=="TET_TKO"]

celltypes <- rbind(foo,bar) %>%
  .[cell%in%sample_metadata$cell] %>%
  .[,c("cell","celltype.mapped","stage.mapped","celltype.score","cellstage.score")]

sample_metadata <- sample_metadata %>% merge(celltypes,all.x=T)
```

# Load data

Load RNA expression data as Seurat object
```{r load_data, echo=FALSE}
seurat <- readRDS(io$seurat)[,sample_metadata$cell]
dim(seurat)
```

Add metadata to the Seurat object
```{r}
foo <- sample_metadata %>% as.data.frame %>% tibble::column_to_rownames("cell") %>%
  .[colnames(seurat),]
stopifnot(colnames(seurat) == rownames(foo))
seurat@meta.data <- foo
# colData(sce) <- foo %>% DataFrame()
```

# Parse data 

Remove genes that are not expressed
```{r}
foo <- Matrix::rowMeans(seurat@assays$RNA@counts)
seurat <- subset(seurat, features = names(which(foo>1e-6)))
```


Normalise using scran, for consistency with the atlas

Convert to SingleCellExperiment
```{r}
sce <- as.SingleCellExperiment(seurat)
rm(seurat)
```

Normalise using the same procedure as in the atlas
```{r}
clusts = as.numeric(quickCluster(sce, method = "igraph", min.size = 100, BPPARAM = mcparam))
```

```{r}
min.clust = min(table(clusts))/2
new_sizes = c(floor(min.clust/3), floor(min.clust/2), floor(min.clust))
sce = computeSumFactors(sce, clusters = clusts, sizes = new_sizes, max.cluster.size = 3000)

to.plot <- data.frame(X = Matrix::colSums(counts(sce)), Y = sizeFactors(sce))
ggplot(to.plot, mapping = aes(x = X, y = Y)) +
  geom_point() +
  labs(x = "Number of UMIs", y = "Size Factor") +
  theme_classic()
```

```{r}
sce <- logNormCounts(sce)
```

Change gene names
```{r}
# rownames(sce) <- rowData(sce)$symbol
```


Load precomputed marker genes
```{r}
marker_genes.dt <- fread(io$atlas.marker_genes)
hvg <- unique(marker_genes.dt$ens_id)
hvg <- hvg[hvg%in%rownames(sce)]

marker_genes.dt[,celltype:=stringr::str_replace_all(celltype,c(" "="_","/"="_"))]
```

Subset SingleCellExperiment to marker genes
```{r}
sce <- sce[hvg,]
```

# Create data.table

Load precomputed average gene expression per cell type from the atlas
```{r}
avg_expr_atlas.dt <- fread(io$atlas.average_expression_per_celltype) %>%
  .[ens_id%in%hvg] %>% 
  setnames("group","celltype")

avg_expr_atlas.dt[,class:="ATLAS"]
avg_expr_atlas.dt[,gene:=NULL]
```

Calculate average gene expression per cell type and genotype in the chimera
```{r}
avg_expr_chimera.dt <- unique(sce$class) %>% map(function(i) {
  sce_filt <- sce[,sce$class==i]
  avg_expr_atlas.dt <- unique(sce_filt$celltype.mapped) %>% map(function(j) {
    dt <- logcounts(sce_filt[,sce_filt$celltype.mapped==j]) %>% as.matrix %>% as.data.table(keep.rownames = T) %>%
      melt(id.vars="rn") %>% setnames(c("ens_id","cell","value")) %>%
      .[,.(mean_expr=round(mean(value),3), detection_rate=round(mean(value>0),3)),by="ens_id"] %>%
      .[,celltype:=j]
    return(dt)
  }) %>% rbindlist %>% .[,class:=i]
}) %>% rbindlist
```

Concatenate
```{r}
avg_expr.dt <- rbind(
  avg_expr_chimera.dt,
  avg_expr_atlas.dt
)
avg_expr.dt[,celltype:=stringr::str_replace_all(celltype,c(" "="_","/"="_"))]
```

# Plot

Compare distribution of expression values
```{r}
gghistogram(avg_expr.dt, x="mean_expr", fill="class", y="..density..") +
  labs(x="Mean expression", y="Density")
```


Consider only lineages with sufficient number of cells
```{r}
opts$min.cells <- 10
foo <- sample_metadata[,.N,by=c("celltype.mapped","class")] %>% 
  setnames("celltype.mapped","celltype") %>%
  .[!is.na(celltype)] %>%
  .[N>=opts$min.cells] %>%
  .[,.N,by="celltype"] %>%
  .[N==2,celltype]

avg_expr.dt <- avg_expr.dt[celltype%in%foo]
```

Scatterplot of atlas expression vs chimera expression per cell type and genotype

```{r}
foo.dt <- sample_metadata[,.N,by=c("celltype.mapped","class")] %>%
  setnames("celltype.mapped","celltype")
```

```{r}
for (i in unique(avg_expr.dt$celltype)) {
  
  to.plot <- avg_expr.dt[celltype==i] %>%
    dcast(ens_id~class,value.var="mean_expr") %>%
    melt(id.vars=c("ens_id","ATLAS"), variable.name="class", value.name="mean_expr") %>%
    setorder(-mean_expr)
  
  # Add (N=) to the genotypes
  to.rename <- c(
    "WT" = sprintf("WT (N=%s)",foo.dt[class=="WT" & celltype==i,N]),
    "TET_TKO" = sprintf("TET_TKO (N=%s)",foo.dt[class=="TET_TKO" & celltype==i,N])
  )
  to.plot[,class:=stringr::str_replace_all(class,to.rename)] 
  
  # Subset lineage-specific genes
  to.plot <- to.plot[ens_id %in% marker_genes.dt[celltype==i,ens_id]]

  p <- ggplot(to.plot, aes(x=ATLAS, y=mean_expr)) +
    # ggrastr::geom_point_rast(color="black", shape=21, size=2, stroke=0.5) +
    # geom_point(color="black", shape=21, size=2, stroke=0.5) +
    scattermore::geom_scattermore(pointsize=3, pixels=c(1000,1000), interpolate=TRUE) +
    stat_smooth(method="lm", color="black", alpha=0.5) +
    stat_cor(method = "pearson") +

    facet_wrap(~class) +
    labs(x="RNA expression atlas", y="RNA expression query") +
    theme_classic()
  
  pdf(sprintf("%s/%s_%s.pdf",io$outdir,i,paste(opts$stages,collapse="-")), width = 8, height = 5)
  print(p)
  dev.off()
}
```

```{r}
  
  # genes.to.label <- marker_genes.dt[celltype==i] %>%
  #   setorder(-score) %>%
  #   head(n=10) %>% .$ens_id
  # to.label <- to.plot[ens_id%in%genes.to.label]

    # ggrepel::geom_text_repel(aes(label=ens_id), size=4, data=to.label) +  
```

